---
title: "Summary Statistics Using tidyverse"
author: "Sharif Islam"
date: '2020-12-29'
output: html_document
---

<style type="text/css">

h1.title {
  font-size: 38px;
  color: DarkRed;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlack;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
</style>

```{r}
library(tidyverse)
```

For illustration purpose, `mtcars` dataset is being used. To create summary statistics, `summarise_each` function is used along with `funs`function. 

```{r}
df <- as_tibble(mtcars)
df.sum <- df %>%
  select(mpg, cyl, vs, am, gear, carb) %>% # select variables to summarise
  summarise_each(funs(Min = min, 
                      Q25 = quantile (., 0.25), 
                      Median = median, 
                      Q75 = quantile (., 0.75), 
                      Max = max,
                      Mean = mean, 
                      StdDev = sd,
                      N = n())) 
```

To organize the raw summary into a nice table, two excellent functions from `tidyverse` - `pivot_longer` and `pivot_wider` - are used. 

```{r}
# Using Pivot_*() function 
df.stats.tidy1 <- df.sum %>% 
  pivot_longer(cols = everything(), names_to = "stat", values_to = "count") %>%
  separate(stat, into = c("Variables", "stat"), sep = "_") %>%
  pivot_wider(names_from = stat, values_from = count) %>%
  select(Variables,N, Mean, StdDev , Min, Q25, Median, Q75, Max )   # reorder columns
```

This is the final summary statistics tables - 

```{r}
df.stats.tidy1
```


Alternatively, we can use the old version of the `pivot_longer` and `pivot_wider`, which are called `gather` and `spread` function respectively. 

```{r }
# Using gather () and spread() function 
df.stats.tidy2 <- df.sum %>% 
  gather(stat, val) %>%
  separate(stat, into = c("Variables", "stat"), sep = "_") %>%
  spread(stat, val) %>%
  select(Variables,N, Mean, StdDev , Min, Q25, Median, Q75, Max ) # reorder columns
```


This is the final summary statistics tables - 

```{r}
df.stats.tidy2
```

